Q Exam Written Report

### 2018-08-09 writeups/q_writeup/q_writeup.tex
final q writeup, my piece of trash thing

\section{Work Accomplished}

@@

        The weakly nonlinear Schr\"odinger equation can be examined for
        \emph{modulational instability} by considering the evolution of modes
        near the propagating plane wave, i.e.\ a wave train centered on
        $\vec{k}_0$. A brief derivation of the criterion is provided
        in~\cite{Whitham} and applied in~\cite{Sutherland1}. The calculations
        show that for certain values of $\vec{k}$, the wave propagates with a
        weakly split group velocity that acts to reduce the amplitude and delay
        breaking, while for other values the group velocity is complex and
        exhibits breaking at lower amplitudes.

@@

            \item[Hyperviscosity] Much more of a numerical trick,
                \emph{hyperviscosity} refers to a family of
                stronger-than-Navier-Stokes viscosities. The one we consider is
                reaching directly into the $k_z$ spectral coefficients and
                artificially decreasing those beyond $N_Z/2$. Conventionally,
                hyperviscosity sets these directly to zero but this seems to be
                too strong by our tests. Because results are still pending, the
                exact hyperviscosity procedure is yet to be determined.
        \end{description}
\end{description}
$N^{-1}$ is the unit of time and $H$ the unit of distance in our computations,
both of which are set to $1$.

>>>>

\section{Work Accomplished: Theory and Numerics}

@@

        The weakly nonlinear Schr\"odinger equation can be examined for
        \emph{modulational instability} by considering the evolution of modes
        near the propagating plane wave, i.e.\ a wave train centered on
        $\vec{k}_0$. A brief derivation of the criterion is provided
        in~\cite{Whitham} and applied in~\cite{Sutherland1}. The calculations
        show that for certain values of $\vec{k}$, the wave propagates with a
        weakly split group velocity that acts to reduce the amplitude and delay
        breaking, while for other values the group velocity is complex and
        exhibits breaking at lower amplitudes. An exmaple of a modulationally
        unstable Gaussian wavepacket is shown in \autoref{fig:modinstab}.
        \begin{figure}[t]
            \centering
            \includegraphics[width=\textwidth]{figs/sutherland_modinstab.png}
            \caption{Modulationally unstable wavepacket breaking before the
            height predicted by linear theory (thick black
            line).}\label{fig:modinstab}
        \end{figure}

@@

            \item[Hyperviscosity] Much more of a numerical trick,
                \emph{hyperviscosity} refers to a family of
                stronger-than-Navier-Stokes viscosities. The one we consider is
                reaching directly into the $k_z$ spectral coefficients and
                artificially decreasing those beyond $N_Z/2$. Conventionally,
                hyperviscosity sets these directly to zero but this seems to be
                too strong by our tests. Because results are still pending, the
                exact hyperviscosity procedure is yet to be determined.
        \end{description}

    \item[Gradual Forcing] One last trick that we are considering is to turn on
        the forcing term gradually; \autoref{fig:plots} shows many examples
        where ringing from turning the forcing term on immediately excites
        ringing that generates spurious nonlinear interactions. We choose to
        effect this by $F(t) \tanh 10\omega_0t$, so over the course of about
        twenty periods the forcing will reach full amplitude. Results are
        pending.
\end{description}

$N^{-1}$ is the unit of time and $H$ the unit of distance in our computations,
both of which are set to $1$. Some snapshots from our simulation can be seen in
\autoref{fig:plots}. Note the smooth
\begin{figure}[t]
    \centering
    \begin{subfigure}{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figs/linear.png}
        \caption{Linear simulation. Note the close agreement of the analytically
        predicted excitement amplitude and the actual waveform. The spectrum is
        broader than a pure Fourier spectrum due to the waveforms being
        sinusoidal and the basis being Chebyshev.}
    \end{subfigure}\hfill
    \begin{subfigure}{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figs/nonlin_psi.png}
        \caption{Nonlinear simulation without regularization. The mismatch of
        the analytical envelope is not significant. Note the significant
        amplitude in higher Chebyshev modes. The simulation breaks soon after
        this point, seemingly due to parametric subharmonic instability.}
    \end{subfigure}

    \begin{subfigure}{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figs/nonlin_ns.png}
        \caption{Nonlinear simulation with Navier-Stokes regularization. The
        mismatch of the analytical envelope is again not significant. The
        supposed PSI from the non-regularized system is clearly just ringing
        from starting the forcing in discontinuous fashion.}
    \end{subfigure}\hfill
    \begin{subfigure}{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figs/nonlin_driving_blowup.png}
        \caption{Another consequence of starting the forcing in discontinuous
        fashion, nonlinear terms amplify the waveform in the driving zone and
        produces nonphysical shocks.}\label{fig:blowup}
    \end{subfigure}
    \caption{The four simulation snapshots above include the following notation:
        the left plot is a simple 2D plot of the $u_{1z}(x, z)$ field, the
        middle plot is a slice $u_{1z}(x=0, z)$ and the right plot plots the
        Chebyshev spectral expansion at the driving $k_x$. In the middle plots,
        the two dashed red lines indicate the damping zones, the dashed blue
        lines indicate the forcing zone $\pm 3\sigma$ and the orange solid lines
        indicate the analytically predicted envelope.}\label{fig:plots}
\end{figure}

\section{Future Work}

Mercifully omitted from this writeup are all of the misguided efforts that were
slowly revised out over time. Nevertheless, against all odds we find ourselves
very near conquering the 2D toy problem, with a few remaining difficulties to be
elaborated below.

\subsection{Remaining 2D Work}

We are very close to having a full working simulation of 2D wave breaking
without numerical singularities. The largest further obstacle that stands in our
way is computational expense, which is being rectified presently within the
group. Compounding our difficulties are that, in the $k_x \ll k_z$ limit that
most astrophysical IGWs find themselves in, the group velocity is extremely
slow, and so requiring resolving $N^{-1}$ timescales results in a very slow
simulation. Even so, we should soon understand where the dissipation in these
excited IGWs occurs as they break via nonlinear processes.

\subsection{3D Work}

\lstinline{Dedalus} has been adapted for 3D work, using spherical harmonics and
Chebyshev polynomials, so the software package is sufficiently up-to-date for
our problem. Nevertheless, 3D calculations will be even more expensive and will
likely carry with them another host of instabilities to sort out. Armed with our
understanding of these 2D difficulties, however, it should be much easier to
tackle.

We also intend to extend our 3D calculations to realstic WD equations of state
and density profiles\cite{fullerI}, such as those in \autoref{fig:eos}.
These will all have to be adapted somewhat to fit the requirements of
\lstinline{Dedalus}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figs/fuller_eos.png}
    \caption{Dynamical variables for a realistic WD model. The thick line is the
    density and the thin line is $N^2$. The dashed line is the square of the
    Lamb frequency, not discussed in this paper but important to 3D
    waves.}\label{fig:eos}
\end{figure}.

\section{Conclusion and Acknowledgements}

In conclusion, we are very close to a numerical solution to the 2D plane
parallel atmosphere IGW breaking problem, with only a few more kinks to sort
out. With this, we should be able to interpret our results within the framework
of existing theory and, after having understood this thoroughly, generalize our
approach to the full realistic 3D problem.

I would like to thank Professor Dong Lai for advising me both as a lab member
and as a first year graduate student during the past year. I would also like to
thank Daniel Lecoanet for explaining his unending back of numerical tricks for
using \lstinline{Dedalus} to me.
